Chord: A Versatile Platform for Program Analysis

### 2011-03-29 main/doc/program_representation.tex
more documentation

\xname{program_representation}
\chapter{Java Program Representation}
\label{chap:program-representation}

Chord uses \xlink{Joeq}{http://joeq.sourceforge.net/} to translate
Java bytecode, one class file at a time, into a three-address-like
intermediate representation of the input Java program called {\it
quadcode}.  This chapter describes how to pretty-print bytecode and
quadcode, which is useful for debugging analyses and deciphering their
output, as well as the API of Joeq and Chord that concerns bytecode
and quadcode, which is useful for users writing their own analyses.

\section{Pretty-Printing Bytecode and Quadcode}

@@

To pretty-print the bytecode of a class, run the following command:
\begin{framed}
\begin{verbatim}
javap -classpath <...> -private -verbose <CLASS NAME>
\end{verbatim}
\end{framed}

where {\tt <CLASS NAME>} is the fully-qualified name of the class
whose bytecode is to be printed (\code{test.HelloWorld} in our
example) and \code{<...>} is the classpath of that class
(\code{examples/hello_world/} in our example).  Program \code{javap}
comes along with the JVM.  The output of the above command for our
example is as follows:

@@

To pretty-print the quadcode of a class, run the following command:

@@

where \code{<...>} is the path of a directory
(\code{examples/hello_world/} in our example) that is expected to
contain a file named \code{chord.properties} which defines
properties \code{chord.main.class} and \code{chord.class.path}
specifying the main class and the classpath of the input Java program
(alternatively, these two properties can be defined directly on the
above command line), {\tt <CLASS NAME>} is the name of the class whose
quadcode is to be printed, and {\tt <...>/<FILE NAME>} is the name of
a file to which the quadcode is to be dumped.  The output of the above
command for our example is as follows:

@@

\section{Traversing Quadcode}

\subsection{Program}

\begin{framed}
\begin{verbatim}
package chord.program;

class Program {
    static Program g() {
        // provides quadcode representation of program
    }
    public IndexSet<jq_Type> getTypes() {
        // provides quadcode representation of all reachable types
    }
    public IndexSet<jq_Reference> getClasses() {
        // provides quadcode representation of all reachable classes
    }
    public IndexSet<jq_Method> getMethods() {
        // provides quadcode representation of all reachable methods
    }
    public Reflect getReflect() {
        // provides information about resolved reflection
    }
    public jq_Reference getClass(String name) {
        // provides quadcode representation of class 
    }
    public jq_Type getType(String name) {
    }
    public jq_Method getMainMethod() {
    }
    public jq_Method getMethod(String mName, String mDesc, String cName) {
    }
    public jq_Method getMethod(String sign) {
    }
    public Quad getQuad(MethodElem e) {
    }
    public Quad getQuad(MethodElem e, Class quadOpClass) {
    }
}    
\end{verbatim}
\end{framed}

\subsection{Type}

\begin{framed}
\begin{verbatim}
class jq_Type {
}
\end{verbatim}
\end{framed}

\subsection{Class}

\begin{framed}
\begin{verbatim}
class jq_Reference {
}
\end{verbatim}
\end{framed}

\subsection{Field}

\begin{framed}
\begin{verbatim}
class jq_Field {
     public String getName();
     public jq_Class getDeclaringClass();
     public jq_Type getType() {
     }
     public String toString() {
         // print this field in format fname:fdesc@cname where:
         // - fname is the name of the field
         // - fdesc is the descriptor of the field's type
         // - cname is the name of the field's declaring class
         // example: buf:[B@java.io.BufferedInputStream
     }
}
\end{verbatim}
\end{framed}

\subsection{Method}

\begin{framed}
\begin{verbatim}
class jq_Method {

}
\end{verbatim}
\end{framed}

Register Factory

Control-Flow Graph

Basic Block

Quad

\begin{framed}
\begin{verbatim}
Traversing a control-flow graph
\end{verbatim}
\end{framed}

\section{Program Descriptors}

MethodSign

@@

\begin{table}
\htmlattributes*{table}{border=1 cellpadding=10}
\begin{tabular}{l|l|l}
BaseType Character	& Type	& Interpretation \\
\hline
{\tt B} &byte& signed byte \\
{\tt C} &char& Unicode character \\
{\tt D} &double& double-precision floating-point value \\
{\tt F} &float& single-precision floating-point value \\
{\tt I} &int& integer \\
{\tt J} &long& long integer \\
{\tt L<classname>;}	&reference& an instance of class {\tt <classname>} \\
{\tt S}	&short& signed short \\
{\tt Z}	&boolean& true or false \\
{\tt [}	&reference& one array dimension
\end{tabular}
\end{table}

@@

A method descriptor represents the parameters that the method takes
and the value that it returns:

@@

A parameter descriptor represents a parameter passed to a method.  A
return descriptor represents the type of the value returned from a
method.  The character \code{V} indicates that the method returns no
value (its return type is void).  For example, the method descriptor
for the method
``\code{Object foo(int i, double d, Thread t)}'' is
``\code{(IDLjava/lang/Thread;)Ljava/lang/Object;}''.
Note that internal forms of the fully qualified names of \code{Thread} and
\code{Object} are used in the method descriptor.  The method descriptor for
\code{foo} is the same whether it is a static or an instance
method. Although an instance method is passed \code{this}, a reference to the
current class instance, in addition to its intended parameters, that
fact is not reflected in the method descriptor.

Chord uniquely represents each bytecode instruction in each method
declared in each class of the program by a
four-tuple \code{bci!mname:mdesc@cname} where \code{bci} is the
(0-based) bytecode offset of the instruction in its containing method,
whose name is \code{mname}, whose descriptor is \code{mname}, and
whose declaring class is \code{cname}.  For
instance, \code{8!main:[Ljava/lang/String;@test.HelloWorld} is the
unique string representation of the \code{return} instruction in the
main method above.

>>>>

\xname{program_representation}
\chapter{Program Representation}
\label{chap:program-representation}

Chord uses \xlink{Joeq}{http://joeq.sourceforge.net/} to translate
Java bytecode, one class file at a time, into a three-address-like
intermediate representation of the input Java program called {\it
quadcode}.  This chapter describes all aspects of quadcode and how it
relates to bytecode.  It first explains how to pretty-print bytecode
and quadcode (Section \ref{sec:pretty-printing}) which is useful for
debugging analyses and deciphering their output.  The remaining
sections describe the quadcode representation along with the API of
Joeq and Chord for navigating it.  Briefly, the representation
consists of a set of classes that may be loaded
(Section \ref{sec:program}).  Each class has a set of fields and
methods, collectively called members of the class
(Section \ref{sec:class-members}).  The representation of a concrete
method (Section \ref{sec:methods}) consists of a control-flow graph
(CFG).  The representation of a CFG (Section \ref{sec:cfgs}) consists
of a set of local variables and a set of basic blocks.  Each basic
block contains zero or more primitive statements called quads
(Section \ref{sec:quads}).  Finally, the most common way to traverse
all quads is discussed (Section \ref{sec:traversing}).

\section{Pretty-Printing}
\label{sec:pretty-printing}

@@

To pretty-print the bytecode representation of a class, run the following command:
\begin{framed}
\begin{verbatim}
javap -classpath <...> -private -verbose <CLASS NAME>
\end{verbatim}
\end{framed}

where:

\begin{itemize}
\item
{\tt <CLASS NAME>} is the fully-qualified name of the class whose
bytecode is to be printed (\code{test.HelloWorld} in our example).
\item
\code{<...>} is the classpath of that class (\code{examples/hello_world/} in our example).
\end{itemize}

Program \code{javap} comes along with the JVM.  The output of the
above command for our example is as follows:

@@

To pretty-print the quadcode representation of a class, run the following command:

@@

where:

\begin{itemize}
\item
\code{<...>} is the path of a directory (\code{examples/hello_world/} in our example) that is expected to
contain a file named \code{chord.properties} which defines
properties \code{chord.main.class} and \code{chord.class.path}
specifying the main class and the classpath of the input Java program
(alternatively, these two properties can be defined directly on the
above command line).
\item
{\tt <CLASS NAME>} is the name of the class whose quadcode is to be
printed.
\item
{\tt <FILE NAME>} is the name of a file to which the quadcode is to be
dumped.
\end{itemize}

The output of the above command for our example is as follows:

@@

\section{Whole Program}
\label{sec:program}

This and the following sections describe the quadcode representation
along with the API of Joeq and Chord for navigating it.
This API is contained in
packages \javadoc{chord.program}{chord/program/package-tree.html}, \javadoc{joeq.Class}{joeq/Class/package-tree.html},
and \javadoc{joeq.Compiler.Quad}{joeq/Compiler/Quad/package-tree.html}.

The quadcode representation of the whole program is a unique global
object of
class \javadoc{chord.program.Program}{chord/program/Program.html}
which can be obtained by calling static
method \javadoc{chord.program.Program.g()}{chord/program/Program.html\#g()}.
This class provides a rich API (in the form of public instance
methods) to access various parts of the representation, most notably:

\begin{mytable}{|l|l|}
\hline
\verb+IndexSet<jq_Type> getTypes()+ & All types referenced in classes that may be loaded. \\
\hline
\verb+IndexSet<jq_Reference> getClasses()+ & All classes that may be loaded.* \\
\hline
\verb+IndexSet<jq_Method> getMethods()+ & All methods that may be called. \T \\
\hline
\end{mytable}

* Includes both classes/interfaces and array types, represented as objects
of \code{jq_Class} and \code{jq_Array}, respectively; both these are
subclasses of \code{jq_Reference}.

See Chapter \ref{chap:computing-scope} for how Chord determines which
classes may be loaded and which methods may be called.

The quadcode representation of each type is a unique object of the
appropriate subclass
of \javadoc{joeq.Class.jq_Type}{joeq/Class/jq_Type.html} in the following
hierarchy:

\begin{framed}
\begin{verbatim}
joeq.Class.jq_Type
  |
  |--- joeq.Class.jq_Primitive
  |
  |--- joeq.Class.jq_Reference
            |
            |--- joeq.Class.jq_Class
            |
            |--- joeq.Class.jq_Array
\end{verbatim}
\end{framed}

\section{Class Members}
\label{sec:class-members}

Each primitive type (e.g., boolean, int, etc.) is represented by a
unique \code{jq_Primtive} object.  Each class and each interface type is
represented by a unique \code{jq_Class} object.  Each array type is
represented by a unique \code{jq_Array} object.

Members (i.e., fields and methods) of the class/interface represented
by an object of
class \javadoc{joeq.Class.jq_Class}{joeq/Class/jq_Class.html} can be
accessed using the following API provided by that class.

\begin{mytable}{|l|p{3.1in}|}
\hline
\verb+String getName()+ & Fully-qualified name of the class, e.g., ``\code{java.lang.String[]}''. \\
\hline
\verb+jq_InstanceField[] getDeclaredInstanceFields()+ & All instance fields declared in the class. \\
\hline
\verb+jq_StaticField[] getDeclaredStaticFields()+ & All static fields declared in the class. \\
\hline
\verb+jq_InstanceMethod[] getDeclaredInstanceMethods()+ & All instance methods declared in the class. \\
\hline
\verb+jq_StaticMethod[] getDeclaredStaticMethods()+ & All static methods declared in the class. \T \\
\hline
\end{mytable}

Chord uses format \code{mName:mDesc@cName}, described in
class \javadoc{chord.program.MethodSign}{chord/program/MethodSign.html},
to uniquely identify each field and each method in the input Java
program, where
\code{mName} denotes the name of the field/method,
\code{mDesc} denotes the descriptor of the field/method (see below),
and \code{cName} denotes the fully-qualified name of the class
declaring the field/method.
For instance, ``\code{main:[Ljava/lang/String;@test.HelloWorld}''
uniquely identifies the main method in the example above.
We next review field descriptors and method descriptors from the Java
bytecode specification.

@@

\begin{mytable}{l|l|l}
\hline
BaseType Character	& Type	& Interpretation \\
\hline
{\tt B} &byte& signed byte \\
{\tt C} &char& Unicode character \\
{\tt D} &double& double-precision floating-point value \\
{\tt F} &float& single-precision floating-point value \\
{\tt I} &int& integer \\
{\tt J} &long& long integer \\
{\tt L<classname>;}	&reference& an instance of class {\tt <classname>} \\
{\tt S}	&short& signed short \\
{\tt Z}	&boolean& true or false \\
{\tt [}	&reference& one array dimension \T \\
\hline
\end{mytable}

@@

A method descriptor represents the types of the arguments and 
return result of a method:

@@

A parameter descriptor represents the type of an argument of a method.
A return descriptor represents the type of the return result of a
method.  The character \code{V} indicates that the method returns no
value (its return type is void).

The method descriptor is the same whether it is a static or an instance
method.  Although an instance method is passed \code{this}, a reference
to the current class instance, in addition to its intended arguments,
that fact is not reflected in the method descriptor.

For example, the method descriptor
for the method ``\code{Object foo(int i, double d, Thread t)}'' is
``\code{(IDLjava/lang/Thread;)Ljava/lang/Object;}''.  Note that
internal forms of the fully qualified names of \code{Thread} and
\code{Object} are used in the method descriptor.

\section{Methods}
\label{sec:methods}

The quadcode representation of each method is a unique object of
class \javadoc{joeq.Class.jq_Method}{joeq/Class/jq_Method.html}.  Components of the method (e.g.,
its control-flow graph) can be accessed using the following API provided
by that class.

\begin{mytable}{|l|p{4in}|}
\hline
\verb+String getName()+ & Name of the method. \\
\hline
\verb+String getDesc().toString()+ & Descriptor of the method, e.g., ``\code{(Ljava/lang/String;)V}''. \\
\hline
\verb+jq_Class getDeclaringClass()+ & Declaring class of the method. \\
\hline
\verb+ControlFlowGraph getCFG()+ & Control-flow graph of the method.* \\
\hline
\verb+int getLineNumber(int bci)+ & Line number of the given bytecode offset (-1 if not found). \\
\hline
\verb+Quad getQuad(int bci)+ & First quad at the given bytecode
offset (null if not found). \\
\hline
\verb+Quad getQuad(int bci, Class kind)+ & First quad of the given
kind at the given bytecode offset (null if not found). \\
\hline
\verb+Quad getQuad(int bci, Class[] kind)+ & First quad of any
given kind at the given bytecode offset (null if not found). \\
\hline
\verb+String toString()+ & Unique identifier of the method in format \code{mName:mDesc@cName}. \T \\
\hline
\end{mytable}

* The control-flow graph must not be asked if the method is abstract (which can be determined by calling instance method \code{isAbstract()} of \code{jq_Method}).

\section{Control-Flow Graphs}
\label{sec:cfgs}

The control-flow graph (CFG) of each method consists of a set of local
variables of the method, called a ``register factory'', and a graph
whose nodes are basic blocks and whose edges denote flow of control
between basic blocks.

The CFG of each method is a unique object of
class \javadoc{joeq.Compiler.Quad.ControlFlowGraph}{joeq/Compiler/Quad/ControlFlowGraph.html}.
Components of the CFG can be accessed using the following API provided
by that class.

\begin{mytable}{|l|l|}
\hline
\verb+getRegisterFactory()+ & Set of all local variables. \\
\hline
\verb+EntryOrExitBasicBlock entry()+ & Unique entry basic block. \\
\hline
\verb+EntryOrExitBasicBlock exit()+ & Unique exit basic block. \\
\hline
\verb+ListIterator.BasicBlock reversePostOrderIterator()+ & Iterator over all basic blocks in reverse post-order. \\
\hline
\verb+jq_Method getMethod()+ & Containing method of the CFG. \T \\
\hline
\end{mytable}

The register factory contains one variable per argument of the method
and one variable per temporary declared in the method body.
Temporaries include those declared by programmers as well as those
generated by Joeq.  The reason for the latter is that the quadcode
representation is register-based and is constructed from Java
bytecode, which is stack-based; moreover, Joeq does the Static Single
Assignment (SSA) transformation by default, which introduces
temporaries to ensure that there is at most one static assignment to
any variable.

The register factory is a unique object
of class \javadoc{joeq.Compiler.Quad.RegisterFactory}{joeq/Compiler/Quad/RegisterFactory.html}.

Each CFG has a set of basic blocks each of which contains a
straight-line sequence of zero or more primitive statements called quads
(Section \ref{sec:quads}).  Each CFG is guaranteed to contain at least
two basic blocks: a unique entry basic block with no incoming edges
and a unique exit block with no outgoing edges.  The entry and exit
basic blocks do not contain any quads.

Each basic block is a unique object of
class \javadoc{joeq.Compiler.Quad.BasicBlock}{joeq/Compiler/Quad/BasicBlock.html}
(the entry and exit basic blocks are instances of a
subclass \javadoc{joeq.Compiler.Quad.EntryOrExitBasicBlock}{joeq/Compiler/Quad/EntryOrExitBsaicBlock.html}).
Components of the basic block can be accessed using the following API
provided by that class.

\begin{mytable}{|l|l|}
\hline
\verb+int size()+ & Number of quads contained in the basic block. \\
\hline
\verb+Quad getQuad(int index)+ & Quad at the given 0-based index. \\
\hline
\verb+List.BasicBlock getPredecessors()+ & List of immediate predecessor basic blocks. \\
\hline
\verb+List.BasicBlock getSuccessors()+ & List of immediate successor basic blocks. \\
\hline
\verb+jq_Method getMethod()+ & Containing method of the basic block. \T \\
\hline
\end{mytable}

\section{Quads}
\label{sec:quads}

Chord uses format \code{offset!mName:mDesc@cName}, described in
class \javadoc{chord.program.MethodElem}{chord/program/MethodElem.html},
to uniquely identify each bytecode instruction in the input Java
program, where \code{offset} is the (0-based) bytecode offset of the
instruction in its containing method, \code{mName} is the name of the
method, \code{mDesc} is the descriptor of the method, and \code{cName}
is the fully-qualified name of the class declaring the method.  For
instance, ``\code{8!main:[Ljava/lang/String;@test.HelloWorld}''
uniquely identifies the return instruction in the main method
in the example above.

The quadcode representation is register-based, as opposed to Java
bytecode that is used to construct it, which is stack-based.  As a
result, it uses {\it quads} to represent bytecode instructions.  A
quad is a primitive statement that consists of an operator and upto
four operands.  There is no one-to-one correspondence between bytecode
instructions and quads: certain bytecode instructions generate a
sequence of more than one quads while others do not generate any quad.
The API of class \code{jq_Method} provides various \code{getQuad(...)}
methods to access the quad(s)
corresponding to a bytecode instruction (see Section \ref{sec:methods}).

Each quad is a unique object of
class \javadoc{joeq.Compiler.Quad.Quad}{joeq/Compiler/Quad/Quad.html}.
Components of the quad can be accessed using the following API
provided by that class.

\begin{mytable}{|l|l|}
\hline
\verb+Operator getOperator()+ & Kind of the quad. \\
\hline
\verb+int getBCI()+ & Bytecode offset of the quad in its containing
method. \\
\hline
\verb+String toByteLocStr()+ & Unique identifier of the quad in
format \code{offset!mName:mDesc@cName}. \\
\hline
\verb+String toJavaLocStr()+ & Location of the quad in
format \code{fileName:lineNum} in Java source code. \\
\hline
\verb+String toLocStr()+ & Location of the quad in both Java bytecode and source code. \\
\hline
\verb+String toVerboseStr()+ & Verbose description of the quad (its location plus contents). \\
\hline
\verb+jq_Method getMethod()+ & Containing method of the quad. \T \\
\hline
\end{mytable}

The kind of each quad is determined by its operator which is a unique object of
the appropriate subclass
of \javadoc{joeq.Compiler.Quad.Operator}{joeq/Compiler/Quad/Operator.html}
in the following hierarchy:

\begin{framed}
\begin{verbatim}
Operator
    |--- Move
    |--- Phi
    |--- Binary
    |--- Unary
    |--- New
    |--- NewAray
    |--- MultiNewArray
    |--- CheckCast
    |--- InstanceOf
    |--- ALength
    |--- ALoad
    |--- AStore
    |--- Getstatic
    |--- Putstatic
    |--- Getfield
    |--- Putfield
    |--- Return
    |--- Branch
    |       |--- IntIfCmp
    |       |--- Goto
    |       |--- Jsr
    |       |--- Ret
    |       |--- LookupSwitch
    |       |--- TableSwitch
    |--- Invoke
    |       |--- InvokeVirtual
    |       |--- InvokeStatic
    |       |--- InvokeInterface
    |--- Monitor
            |--- MONITORENTER
            |--- MONITOREXIT
\end{verbatim}
\end{framed}

The number and kinds of operands of each quad depends upon the kind of
the operator.  Each of the above subclasses of \code{Operator}
provides an API to access the operands of the quad.  For instance,
the components of a \code{Getfield} quad {\tt q} of the form ``l = b.f'' can be accessed as follows:

\begin{framed}
\begin{verbatim}
Operand lo = Getfield.getSrc(q);
Operand bo = Getfield.getBase(q);
if (lo instanceof RegisterOperand && bo instanceof RegisterOperand) {
    Register l = ((RegisterOperand) lo).getRegister();
    Register b = ((RegisterOperand) bo).getRegister();
    jq_Field f = Getfield.getField(q).getField();
    ...
}
\end{verbatim}
\end{framed}

\section{Traversing Quadcode}
\label{sec:traversing}

A common way to traverse all quads in the input Java program is as follows:

\texonly{\newpage}

\begin{framed}
\begin{verbatim}

import chord.program.Program;
import joeq.Compiler.Quad.QuadVisitor;
import joeq.Class.jq_Method;
import joeq.Compiler.Quad.ControlFlowGraph;
import joeq.Util.Templates.ListIterator;
import joeq.Compiler.Quad.BasicBlock;
import joeq.Compiler.Quad.Quad;

QuadVisitor qv = new QuadVisitor.EmptyVisitor() {
    public void visitMove(Quad q) { ... }
    public void visitPhi(Quad q) { ... }
    public void visitBinary(Quad q) { ... }
    ...
};
Program program = Program.g();
for (jq_Method m : program.getMethods()) {
    if (!m.isAbstract()) {
        ControlFlowGraph cfg = m.getCFG();
        ListIterator.BasicBlock it = cfg.reversePostOrderIterator();
        while (it.hasNext()) {
            BasicBlock b = it.nextBasicBlock();
            for (int i = 0; i < b.size(); i++) {
                Quad q = b.getQuad(i);
                q.accept(qv);
            }
        }
    }
}
\end{verbatim}
\end{framed}
